Optimal Estimation with Nuisance Parameters

### 2024-02-19 references/literature/papers.qmd
Added REML papers.

>>>>

# @patterson1971recovery

Title: Recovery of inter-block information when block sizes are unequal

Author(s): H.D. Patterson and R. Thompson

Journal (Year): Biometrika (1971)

Importance: This describes REML for unbalanced designs.

Innovation: They decompose the likelihood into error constrasts and other. (These
are the $S$ and $Q$ matrices.)

# @harville1977maximum

Title: Maximum Likelihood Approaches to Variance Component Estimation and to
Related Problems

Author(s): David A. Harville

Journal (Year): JASA (1977)

Importance: This paper summarizes the use of ML and REML methods and focuses on
their use and computation.

Innovation: This is a good summary of the problems that REML solves.

Title:

Author(s):

Journal (Year):

Importance:

Innovation:

Title:

Author(s):

Journal (Year):

Importance:

Innovation:

Title:

Author(s):

Journal (Year):

Importance:

Innovation:

### 2024-02-19 src/explore/20240216-optest.qmd
Added REML papers.

The goal of the simulation is to estimate $\theta = E[Y]$. We also estimate
several nuisance parameters $\eta_1 = E[X_1]$ and $\eta_2 = E[X_2]$. All of
these parameters can be estimated using GLS estimation. The model that we
propose is the following. Consider the model, for 
$g = (\eta_1, \eta_2, \theta)'$

@@

However, we can do better if we assume that the distribution of $\hat g$ is a
joint multivariate normal model and compute $\hat \theta \mid \hat \eta_1, \hat
\eta_2$. We can represent this model^[The "C" is for conditional.] as 
$\hat \theta_C := E[\hat \theta \mid \hat \eta]$. For each of these models we
use the true covariance matrix and the estimated covariance matrix. For a full
discussion of the derivation of $\hat \theta_C$ see Appendix A.

@@

@@

>>>>

The goal of the simulation is to estimate $\theta = E[Y]$. We also estimate
several nuisance parameters $\eta_1 = E[X_1]$ and $\eta_2 = E[X_2]$. All of
these parameters can be estimated using GLS estimation.

## The Marginal GLS Model

The model that we propose is the following. Consider the model, for 
$g = (\eta_1, \eta_2, \theta)'$

@@

## The Conditional GLS Model

However, we can do better if we assume that the distribution of $\hat g$ is a
joint multivariate normal model and compute $\hat \theta \mid \hat \eta_1, \hat
\eta_2$. We can represent this model^[The "C" is for conditional.] as 
$\hat \theta_C := E[\hat \theta \mid \hat \eta]$. In this representation, 
the initial goal is to minimize $Q(\eta_1, \eta_2, \theta)$ where

More details about how we solve this model in in Appendix A.
For each of these models we use the true covariance matrix and the estimated
covariance matrix.

@@

Comment on the second bullet point. Currently, I estimate the covariance matrix
by estimating the corresponding expected values separately using all of the
available data. See Appendix B for more details. For example, to estimate 
$E[X_1 X_2]$ we use

$$\hat E[X_1 X_2] = 
\frac{\sum_{i = 1}^n (\delta_{10} + \delta_{11})(x_{1i}x_{2i})}{\sum_{i = 1}^n
\delta_{10} + \delta_{11}}.
$$

I think that this may be overcome using REML. I will be working on this.

@@

# Appendix B: Computing the Conditional Covariance Matrix

For the simulation study in this report the covariance matrix for the
conditional model $\hat \theta_C$. Since each segment is considered to be
independent, using the current estimator $\hat \theta_C$ we only need to compute
the variance of each $\hat f_j$ and the following covariances: 
$\Cov(\hat f_1^{(11)}, \hat f_2^{(11)})$,
$\Cov(\hat f_1^{(11)}, \hat f_3^{(11)})$,
$\Cov(\hat f_2^{(11)}, \hat f_3^{(11)})$,
$\Cov(\hat f_1^{(10)}, \hat f_2^{(10)})$, and 
$\Cov(\hat f_1^{(01)}, \hat f_3^{(01)})$.

The variances of each $\hat f_j$ are relatively straightforward. As an example,
I will derive $\Var(\hat f_1^{(11)})$.

$$
\Var(\hat f_1^{(11)}) = n^{-1}(\pi_{11}^{-1} E[X_1^2] - E[X_1]^2).
$$

For other values of $\hat f_j$ we simply use the correct $\pi_{ij}$ and variable
from $Z = (X_1, X_2, Y)$. In a similar manner, the covariance 
$\Cov(\hat f_1^{(11)}, \hat f_2^{(11)})$, can be computed as

$$
\Cov(\hat f_1^{(11)}, \hat f_2^{(11)}) =
n^{-1}(\pi_{11}^{-1} E[X_1 X_2] - E[X_1]E[X_2]).
$$

For the other covariances, we can substitute the corresponding $\pi_{ij}$ and
other variables from $Z$.

We using the true covariance, I use the population covariance that I know from
the construction of the problem. To estimate the covariance matrix, I estimate
each required expectation (such as $E[X_1 X_2]$, $E[X_1]$, and $E[X_2]$ for the
previous covariance) using all of the available data. This means that

$$
\hat E[X_1 X_2] = \frac{\sum_{i = 1}^n (\delta_{11} + \delta_{10})x_{1i}x_{2i}}{
\sum_{i = 1}^n (\delta_{11} + \delta_{10})},
\hat E[X_1] = n^{-1} \sum_{i = 1}^n x_{1i} \text{ and } 
\hat E[X_2] = \frac{\sum_{i = 1}^n (\delta_{01} + \delta_{11})x_{2i}}{\sum_{i =
1}^n \delta_{01} + \delta_{11}}.
$$
